---
output: html_document
---
Exploration of Red Wines Quality by Yuwen Tan
========================================================

This report explores the dataset of the qulity of red wines. The dataset is downloaded from "Data Set Options" document on Udacity. We would look into the dataset and try to find how the quality of red wines is affected by different features.

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```

```{r, load packages}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

library(ggplot2)
library(gridExtra)
library(psych)
library(rpart)
library(rpart.plot)
```

```{r, "load the data"}
# Load the Data
redwines <- read.csv('wineQualityReds.csv')

# remove X because it only represents the row numbers
redwines$X <- NULL
```

# Univariate Plots Section

```{r, "dataset dimension"}
dim(redwines)
```

The dataset consists of 12 variables and 1599 observations.

However, some records are duplicated. For example:
```{r, "show duplicated records"}
redwines[244:245,]
```

I don't think two wines would have eactly the same feature values. It is reasonable to consider the duplicate records were mistakenly recorded twice or more, and only one of them should be kept in our dataset. I looked into the data to find all duplicate records and removed redundant ones:

```{r, "find duplicate records and keep only one of them"}
# find duplicate record id
duplicate_ind <- duplicated(redwines)

# count total duplicates
sum(duplicate_ind)

# remove duplicates
redwines <- redwines[!duplicate_ind,]
```

There are 240 duplicates that we removed from the dataset. Now we can summarize our dataset again:

```{r, "cleaned dataset summary"}
dim(redwines)
str(redwines)
summary(redwines)
```

The cleaned dataset now consists of 12 variables and 1359 observations. All variables are of numeric values.

```{r, "keep a factor copy of quality"}
# copy quality as factor, to make plotting easier in later analyses
redwines$quality.factor <- as.factor(redwines$quality)
```

```{r, "univariate plot: quality"}
ggplot(aes(x = quality), data = redwines) + 
  geom_histogram(binwidth = 0.5) +
  scale_x_continuous(name = 'quality', breaks = seq(3,8))

# show quality counts from high to low
sort(table(redwines$quality), decreasing = TRUE)[1:6]
```

The variable of "quality" in this dataset only takes value of a integer ranging from 3 to 8. From the description of the dataset, quality actually could be a integer between 0 and 10. 0 stands for very bad quality, and 10 stands for very excellent quality. In this dataset, most values of quality are 5 or 6; very few of the observations (only 17) have the highest quality of 8.

```{r, "univariate plot: fixed.acidity"}
ggplot(aes(x = fixed.acidity), data = redwines) + 
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(name = 'fixed acidity', breaks = seq(4,16))

summary(redwines$fixed.acidity)
```

The variable of "fixed acidity" ranges from 4.6 to 15.9. Most values of fixed acidity are within range from 6 to 10.

```{r, "univariate plot: volatile.acidity"}
ggplot(aes(x = volatile.acidity), data = redwines) + 
  geom_histogram(binwidth = 0.01) +
  scale_x_continuous(name = 'volatile acidity', breaks = seq(0,1.6,0.1))

summary(redwines$volatile.acidity)
```

The variable of "volatile acidity" ranges from 0.12 to 1.58. The distribution of volatile acidity seems bimodal with peaks around 0.4 and 0.6. The highest value of 1.58 only appears in one observation and is much higher than majority of the values. We can take a look at this data point:

```{r, "special point: volatile.acidity"}
redwines[redwines$volatile.acidity==1.58,]
```

This seems to be a valid record, with the lowest quality value of 3.

```{r, "summary: volatile.acidity of different quality level"}
summary(subset(redwines$volatile.acidity,redwines$quality <= 4))
summary(subset(redwines$volatile.acidity,redwines$quality >= 5 && redwines$quality <= 6))
summary(subset(redwines$volatile.acidity,redwines$quality >= 7))
```

Summaries above shows the volatile acidity of wines with low quality (3/4), medium quality (5/6), and high quality (7/8) respectively. It shows that low quality wines have a median volatile acidity of 0.68, higher than the third quantile volatile acidity of 0.64 for medium quality wines. Similarly, medium quality wines have a median volatile acidity of 0.52, higher than the third quantile volatile acidity of 0.4825 for high quality wines.

```{r, "plot by quality level: volatile acidity of different quality level"}
ggplot(aes(x = volatile.acidity), data = redwines) +
  facet_wrap(~quality, nrow = 3, ncol = 2, scales = 'free_y') +
  geom_histogram(binwidth = 0.05)
```

Above are the histograms of volatile acidity for different qualities. We can clearly the peak in each plot is moving from high to low as volatile acidity decreases.

```{r, "univariate plot: citric.acidity"}
ggplot(aes(x = citric.acid), data = redwines) + 
  geom_histogram(binwidth = 0.01) +
  scale_x_continuous(name = 'citric acid', breaks = seq(0,1,0.1))

summary(redwines$citric.acid)

# show citric acid values with top 10 counts
sort(table(redwines$citric.acid), decreasing = TRUE)[1:10]
```

The variable of "citric acid" ranges from 0 to 1. The distribution of citric acid seems to be very random with several peaks like 0, 0.49, 0.24, and 0.02. The highest value of 1 only appears in one observation and is much higher than majority of the values. We can take a look at this data point:

```{r, "special point: citric.acid"}
redwines[redwines$citric.acid==1,]
```

This seems to be a valid record, with a low quality value of 4.

```{r, "univariate plot: residual.sugar"}
ggplot(aes(x = residual.sugar), data = redwines) + 
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(name = 'residual sugar', breaks = seq(0,16,1))

summary(redwines$residual.sugar)
```

The variable of "residual sugar" ranges from 0.9 to 15.5. Most values of residual sugar are around 2. The lowest and highest points are displayed below:

```{r, "special point: residual.sugar"}
redwines[redwines$residual.sugar < 1 | redwines$residual.sugar > 15,]
```

They all seem to be valid records. 

```{r, "univariate plot: chlorides"}
ggplot(aes(x = chlorides), data = redwines) + 
  geom_histogram(binwidth = 0.01) +
  scale_x_continuous(name = 'chlorides', breaks = seq(0,0.7,0.05))

summary(redwines$chlorides)
```

The variable of "chlorides" ranges from 0.012 to 0.611. Most values of chlorides are around 0.075. The lowest and highest points are displayed below:

```{r, "special point: chlorides"}
redwines[redwines$chlorides < 0.02 | redwines$chlorides > 0.6,]
```


```{r, "univariate plot: free.sulfur.dioxide"}
ggplot(aes(x = free.sulfur.dioxide), data = redwines) + 
  geom_histogram(binwidth = 1) +
  scale_x_continuous(name = 'free sulfur dioxide', breaks = seq(0,75,5))

summary(redwines$free.sulfur.dioxide)
```

The variable of "free sulfur dioxide" ranges from 1 to 72. Most values of free sulfur dioxide are around 5. The lowest and highest points are displayed below:

```{r, "special point: free.sulfur.dioxide"}
redwines[redwines$free.sulfur.dioxide <= 1 | redwines$free.sulfur.dioxide >= 70,]
```

```{r, "univariate plot: total.sulfur.dioxide"}
ggplot(aes(x = total.sulfur.dioxide), data = redwines) + 
  geom_histogram(binwidth = 2) +
  scale_x_continuous(name = 'total sulfur dioxide', breaks = seq(0,300,50))

summary(redwines$total.sulfur.dioxide)
```

The variable of "total sulfur dioxide" ranges from 6 to 289. Most values of total sulfur dioxide are around 25. The highest two points are much higher than the rest of the points. They are displayed below:

```{r, "special point: total.sulfur.dioxide"}
redwines[redwines$total.sulfur.dioxide >= 250,]
```

Interestingly, their values of features other than total sulfur dioxide are excatly the same. However, we don't have enought information to decide how these two records are related. We would keep both of them in the dataset.

```{r, "univariate plot: density"}
ggplot(aes(x = density), data = redwines) + 
  geom_histogram(binwidth = 0.0001) +
  scale_x_continuous(name = 'density', breaks = seq(0.99,1.004,0.002))

summary(redwines$density)
```

The variable of "density" ranges from 0.9901 to 1.004. The distribution of desity values seems to be normal, centered around 0.997.

```{r, "univariate plot: pH"}
ggplot(aes(x = pH), data = redwines) + 
  geom_histogram(binwidth = 0.01) +
  scale_x_continuous(name = 'pH', breaks = seq(2.7,4.1,0.1))

summary(redwines$pH)
```

The variable of "pH" ranges from 2.74 to 4.01. The distribution of pH values seems to be normal, centered around 3.3.

```{r, "univariate plot: sulphates"}
ggplot(aes(x = sulphates), data = redwines) + 
  geom_histogram(binwidth = 0.01) +
  scale_x_continuous(name = 'sulphates', breaks = seq(0,2,0.2))

summary(redwines$sulphates)
```

The variable of "sulphates" ranges from 0.33 to 2. Most values of sulphate are within range from 0.4 to 0.8. We could take another look at the variable by zooming in and applying a log transformation to the scale of x axis:

```{r, "univariate plot: sulphates zoom in and log tranform x axis"}
ggplot(aes(x = sulphates), data = subset(redwines, sulphates < 1.2)) + 
  geom_histogram(binwidth = 0.005) +
  scale_x_log10(name = 'sulphates', breaks = seq(0,2,0.2))
```

Now the distribution of sulphates seems to be normal in the newly scaled plot.

```{r, "summary: sulphates of different quality level"}
summary(subset(redwines$sulphates,redwines$quality <= 4))
summary(subset(redwines$sulphates,redwines$quality >= 5 && redwines$quality <= 6))
summary(subset(redwines$sulphates,redwines$quality >= 7))
```

Summaries above shows the sulphates of wines with low quality (3/4), medium quality (5/6), and high quality (7/8) respectively. It shows that low quality wines have median sulphates of 0.56, close to the first quantile sulphates of 0.55 for medium quality wines. Similarly, medium quality wines have median sulphates of 0.62, lower than the first quantile sulphates of 0.6575 for high quality wines.

```{r, "plot by quality level: sulphates of different quality level"}
ggplot(aes(x = sulphates), data = redwines) +
  facet_wrap(~quality, nrow = 3, ncol = 2, scales = 'free_y') +
  geom_histogram(binwidth = 0.05)
```

Above are the histograms of sulphates for different qualities. We can clearly the peak in each plot is moving from low to high as sulphates increase.

```{r, "univariate plot: alcohol"}
ggplot(aes(x = alcohol), data = redwines) + 
  geom_histogram(binwidth = 0.05) +
  scale_x_continuous(name = 'alcohol', breaks = seq(8,15,1))

summary(redwines$alcohol)
```

The variable of "alcohol" ranges from 8.4 to 14.9. Most values of alcohol are within range from 9 to 13.

```{r, "special point: alcohol"}
redwines[redwines$alcohol < 9 | redwines$alcohol > 14,]
```

The highest and lowest observations of alcohol all seem to be valid.

# Univariate Analysis

### What is the structure of your dataset?

There are 1599 observations of wines in the dataset with 12 features. However, some of them are completely identical. After removing duplicates, there are different 1359 observations.

The 12 features are as follows: fixed acidity, volatile acidity, citric acid, residual sugar, chlorides, free sulfur dioxide, total sulfur dioxide, density, pH, sulphates, alcohol, and quality.

Quality only takes value of a integer ranging from 3 to 8 in the dataset, and most wines in the dataset have the quality of 5 or 6. We can consider quality as a numeric variable since the higher the quality the better the wine.

### What is/are the main feature(s) of interest in your dataset?

The main features I am interested in are volatile acidity and sulphates. They have relatively different value ranges for different qualities. We may build a predictive model based on these features. It would be fun to predict the subjective quality from objective features of a wine.

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

Other features that might help are alcohol, fixed acidity, and citric acid. It is natural to think that if the alcohol is too low the wine would not get a high score from tasters. Given that volatile acidity is considered as one of the main features, it would be helpful to include the other two kind of acidity in our study since they may be somehow correlated.

### Did you create any new variables from existing variables in the dataset?

No new variables are created in the dataset.

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?

There are some observations I considered as duplicates and they are removed from the original dataset. I think it is unlikely to have two or more records exactly the same in the dataset, because many features are so detailed with precision to 0.0001. Therefore, I removed duplicates to avoid them from affecting the analysis.

I also log-transformed the right-skewed feature of sulphates. The transformed distribution appears to be normal and peaking around 0.6.

# Bivariate Plots Section
```{r, "correlations"}
# show the relationship between features we selected
part1 = c('fixed.acidity', 'volatile.acidity', 'citric.acid',
          'sulphates', 'alcohol', 'quality')
redwines.part1 = redwines[, part1]
cor(redwines.part1)
pairs.panels(redwines.part1)

# show the relationship between features we did not select
part2 = c('residual.sugar', 'chlorides', 'free.sulfur.dioxide',
          'total.sulfur.dioxide', 'density', 'pH', 'quality')
redwines.part2 = redwines[, part2]
cor(redwines.part2)
pairs.panels(redwines.part2)
```

From the correlation matrix and pair plot above, I grouped the features into tow groups. The first part shows whatever features we are interested and their correlations with quality. The second part shows rest of the features and their correlations with quality. It is clear that features in the first group show stronger correlations with quality.

Among all features, alcohol is most positively correlated with quality, and volatile acidity is most negatively correlated with quality. However, the absolute value of those two correlations are less than 0.5, which do not indicate a strong correlation. We will take a closer look.

```{r, "volatile.acidity: boxplot"}
ggplot(redwines, aes(quality, volatile.acidity)) +
  geom_boxplot(aes(group=quality), outlier.shape = NA) +
  geom_jitter(alpha = 0.2, position = position_jitter(width = 0.4)) +
  scale_x_continuous(name = 'quality', breaks = seq(3,8,1)) +
  scale_y_continuous(name = 'volatile acidity')
```

The boxplot above clearly shows that as quality increases, volatile acidity decreases.

```{r, "volatile.acidity: linear regression"}
# apply linear model to see the relationship betwwen quality and volatile.acidity
lm.volatile.acidity = lm(formula = quality ~ volatile.acidity, data = redwines)

summary(lm.volatile.acidity)
```

The linear regression shows that volatile acidity is significant if we fit a linear model for quality. However, the R-squared is only around 0.16, which means that such a linear model is not doing so well.

```{r, "sulphates: boxplot"}
ggplot(redwines, aes(quality, sulphates)) +
  geom_boxplot(aes(group=quality), outlier.shape = NA) +
  geom_jitter(alpha = 0.2, position = position_jitter(width = 0.4)) +
  scale_x_continuous(name = 'quality', breaks = seq(3,8,1)) +
  scale_y_continuous(name = 'sulphates', breaks = seq(0,2,0.2))
```

The boxplot above shows that as quality increases, sulphates increase. One takeaway is that, most values of sulphates are less than 0.9, no matter what the quality is.

```{r, "sulphates: linear regression"}
# apply linear model to see the relationship betwwen quality and sulphates
lm.sulphates = lm(formula = quality ~ sulphates, data = redwines)

summary(lm.sulphates)
```

The linear regression shows that the feature of sulphates is significant if we fit a linear model for quality. However, the R-squared is only around 0.06. The model is not good (or say "very bad"). We may think what if we seperate the data into two parts: sulphates less than 0.9 and sulphates more than or equal to 0.9.

```{r, "sulphates: boxplot - two parts"}
plot1 <- ggplot(subset(redwines, sulphates < 0.9), aes(quality, sulphates)) +
  geom_boxplot(aes(group=quality), outlier.shape = NA) +
  geom_jitter(alpha = 0.2, position = position_jitter(width = 0.4)) +
  scale_x_continuous(name = 'quality', breaks = seq(3,8,1)) +
  scale_y_continuous(name = 'sulphates') +
  ggtitle('sulphates less than 0.9')

plot2 <- ggplot(subset(redwines, sulphates >= 0.9), aes(quality, sulphates)) +
  geom_boxplot(aes(group=quality), outlier.shape = NA) +
  geom_jitter(alpha = 0.2, position = position_jitter(width = 0.4)) +
  scale_x_continuous(name = 'quality', breaks = seq(3,8,1)) +
  scale_y_continuous(name = 'sulphates') +
  ggtitle('sulphates more than or equal to 0.9')

grid.arrange(plot1, plot2)
```

The two boxplots above show a very interesting trend. For sulphates less than 0.9, as quality increases, sulphates increase. However, for sulphates more than or equal to 0.9, the trend is somehow the other way around. This actually makes some sense: wines of the best quality might be balanced in some features, not being too high or too low.

```{r, "sulphates: linear regression - sulphates lower than 0.9"}
# apply linear model to see the relationship betwwen quality and sulphates
# this time only test on those with sulphates lower than 0.9
lm.sulphates.low = lm(formula = quality ~ sulphates, data = subset(redwines, sulphates < 0.9))

summary(lm.sulphates.low)
```

The linear regression above shows that if we only look into observations with sulphates less than 0.9, the feature of sulphates is still significant if we fit a linear model for quality. Moreover, the R-squared increases to around 0.18.

```{r, "alcohol: boxplot"}
ggplot(redwines, aes(quality, alcohol)) +
  geom_boxplot(aes(group=quality), outlier.shape = NA) +
  geom_jitter(alpha = 0.2, position = position_jitter(width = 0.4)) +
  scale_x_continuous(name = 'quality', breaks = seq(3,8,1)) +
  scale_y_continuous(name = 'alcohol', breaks = seq(8,15,1))
```

The boxplot above shows that except quality value of 5, as quality increases, alcohol increases. For quality value of 5, there are too many observations with alcohol between 9 to 9.5, much frequent than other quality values.

```{r, "alcohol: linear regression"}
# apply linear model to see the relationship betwwen quality and alcohol
lm.alcohol = lm(formula = quality ~ alcohol, data = redwines)

summary(lm.alcohol)
```

The linear regression shows that alcohol is significant if we fit a linear model for quality. However, the R-squared is only around 0.23. The model is not good enough, but better than the previous models.

```{r, "scatterplot: volatile.acidity and fixed acidity"}
ggplot(aes(x = volatile.acidity, y = fixed.acidity), data = redwines) +
  geom_point(alpha = 0.2) + 
  scale_x_continuous(name = 'volatile acidity') +
  scale_y_continuous(name = 'fixed acidity')
```

The scatterplot shows that there is no strong relationship between volatile acidity and fixed acidity. After looking up information about wine, I found that the total acidity in wine is usually divided into two groups: volatile and fixed. Therefore, it is understandable that volatile acidity and fixed acidity are not so related.

```{r, "scatterplot: citric acid compared to other acidity"}
plot1 <- ggplot(aes(x = citric.acid, y = fixed.acidity), data = redwines) +
  geom_point(alpha = 0.2) + 
  scale_x_continuous(name = 'citric acid') +
  scale_y_continuous(name = 'fixed acidity')

plot2 <- ggplot(aes(x = citric.acid, y = volatile.acidity), data = redwines) +
  geom_point(alpha = 0.2) + 
  scale_x_continuous(name = 'citric acid') +
  scale_y_continuous(name = 'volatile acidity')

grid.arrange(plot1, plot2)
```

The scatterplots above actually show that citric acid is clearly positively correlated with fixed acidity, and somehow negatively correlated with volatile acidity. This is actully quite interesting, since citric is one of the predominant fixed acidity found in wines. Although fixed acidity is not correlated with volatile acidity, citric acid is correlated with both of them.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

Among all features, alcohol and volatile acidity have the strongest correlation with quality. The correlations are 0.48 and -0.39 respectively. Other features do not show strong correlation with quality, with absolute correlations lower than 0.25.

However, there is not a simple feature that could explain quality very well. Based on the R-square values, alcohol or volatile can only explain around 20% of the variance in quality.

Interestingly, the feature of sulphates has much better predictive power after we sepearte the dataset into to two parts according to the value of sulphates. This introduces me a new way to look into the dataset. We can have more investigation about this phenomenon in the multivariate plots section.

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

Fixed acidity and volatile acidity are almost not correlated in the dataset. This might be because that such the classification of "fixed" and "volatile" is systematically designed according to chemical characters, and these two types are kind of independent within the classification.

Another interesting finding is about citric acid. It is supposed to be one kind of fixed acidity, and we do see that it is highly positively correlated with fixed acidity. However, it is also somehow negatively correlated with volatile acidity.

### What was the strongest relationship you found?

The strongest features related to quality are alcohol and volatile acidity. The feature of sulphates may be useful in building the model to predict quality, but it need to be carefully grouped before fitting.

# Multivariate Plots Section

```{r, "volatile.acidity & fixed.acidity vs citric.acid"}
ggplot(aes(x = volatile.acidity, y = fixed.acidity, color = citric.acid), 
       data = redwines) +
  geom_point(alpha = 0.5, size = 1) +
  scale_x_continuous(name = 'volatile acidity') +
  scale_y_continuous(name = 'fixed acidity') +
  scale_color_gradient(low="blue", high="red") +
  labs(colour = "citric acid") +
  ggtitle("Citric Acid against Volatile and Fixed Acidity")
```

This plot shows citric acid against volatile acidity vs. fixed acidity relationship. This helps to explain why we see little correlation between volatile acidity and fixed acidity, but higher correlation between them and citric acid. Wines with high citric acid tend to stay in the top left in the plot, while wines with low citric acid tend to stay in the bottom right in the plot.

```{r, "volatile.acidity & alcohol vs quality"}
ggplot(aes(x = volatile.acidity, y = alcohol, color = quality.factor), data = redwines) +
  geom_point(alpha = 0.5, size = 1) +
  scale_x_continuous(name = 'volatile acidity') +
  scale_y_continuous(name = 'alcohol') +
  scale_color_brewer(palette="YlOrRd") +
  labs(colour = "quality") +
  ggtitle("Quality against Volatile Acidity and Alcohol") +
  theme_dark()
```

In this plot, I showed the quality against volatile acidity vs. alcohol relationship. It seems that we could more easily distinguish quality of 5 and 7 by checking the value of alcohol (use alcohol value of 10 as the decision boundary), while we could more easily distinguish quality of 6 and 7 by checking the value of volatile acidity (use volatile acidity value of 0.4 as the decision boundary).

```{r, "sulphates & alcohol vs quality"}
ggplot(aes(x = sulphates, y = alcohol, color = quality.factor), data = redwines) +
  geom_point(alpha = 0.5, size = 1) +
  scale_x_continuous(name = 'sulphates') +
  scale_y_continuous(name = 'alcohol') +
  scale_color_brewer(palette="YlOrRd") +
  labs(colour = "quality") +
  ggtitle("Quality against Sulphates and Alcohol") +
  theme_dark()
```

In this plot, I showed the quality against sulphates vs. alcohol relationship, similar to what we have just shown. It seems that besides that we could still distinguish quality of 5 and 7 by checking the value of alcohol (use alcohol value of 10 as the decision boundary), we can also distinguish quality of 5 and 7 by checking the value of sulphates (use sulphates value of 0.7 as the decision boundary). These two features combined help us better distinguish quality of 5 or 7.

At the same time, we could more easily distinguish quality of 6 and 7 by checking the value of sulphates (use sulphates value of 0.75 as the decision boundary). It would be very hard to distinguish the quality of 6 and 7 by checking the value of alcohol.

```{r, "sulphates & volatile.acidity vs quality"}
ggplot(aes(x = sulphates, y = volatile.acidity, color = quality.factor), data = redwines) +
  geom_point(alpha = 0.5, size = 1) +
  scale_x_continuous(name = 'sulphates') +
  scale_y_continuous(name = 'volatile acidity') +
  scale_color_brewer(palette="YlOrRd") +
  labs(colour = "quality") +
  ggtitle("Quality against Sulphates and Volatile Acidity") +
  theme_dark()
```

In this plot, I showed the quality against sulphates vs. volatile acidity relationship. The plot shows that these two features combined help to distinguish quality of 6 or 7. Wines with quality of 6 tend to stay in the top left in the plot, while wines with quality 7 tend to stay in the bottom right in the plot.

```{r, "decision tree: volatile.acidity, sulphates, alcohol"}
# apply tree model to try predict quality
output.tree <- rpart(quality.factor ~ volatile.acidity + alcohol + sulphates,
                     data = redwines, method = 'class',
                     control = rpart.control(minsplit=100, cp=0))
prp(output.tree)

# calculate the fitting accuracy of the tree model
accuracy.table = table(predict(output.tree, redwines, type = 'class'), 
                       redwines$quality.factor)
colnames(accuracy.table) = c('Forecast # of Quality 3',
                             'Forecast # of Quality 4',
                             'Forecast # of Quality 5',
                             'Forecast # of Quality 6',
                             'Forecast # of Quality 7',
                             'Forecast # of Quality 8')
rownames(accuracy.table) = c('Actual # of Quality 3',
                             'Actual # of Quality 4',
                             'Actual # of Quality 5',
                             'Actual # of Quality 6',
                             'Actual # of Quality 7',
                             'Actual # of Quality 8')
accuracy.table
correct.counts = 0
for (i in 1:6){
  correct.counts = correct.counts + accuracy.table[i,i]
}
accuracy.rate = correct.counts/dim(redwines)[1]
cat("Accuray: ", accuracy.rate)
```

I created a decision tree model to predict quality based on volatile acidity, alcohol, and sulphates. It is actually doing well. Introduicing more features may increase the accuracy.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

I do find that some varibles could be used to distinguish some quality values, while other variables could used to distinguish other quality values. For example, to distinguish quality of 5 and 7, it is better to check the values of alcohol and sulphates; to distinguish quality of 6 and 7, it is better to check the value of sulphates and volatile acidity. Some features combined togetehr would help to distinguish the value of quality, like I just listed.

### Were there any interesting or surprising interactions between features?

The interactions between citric acid and two types of acidity (fixed and volatile) are interesting. Wines with higher citric acid are more likely to be high in fixed acidity and low in volatile acidity, and wines with lower citric acid are more likely to be low in fixed acidity and high in volatile acidity. This explains why citric acid is correlated with two types of acidity while two types of acidity could be uncorrelated.

### OPTIONAL: Did you create any models with your dataset? Discuss the strengths and limitations of your model.

I created a decision tree model. This model is simple since it only uses three features. It also has good predicting power, much better than simple guessing. This model also makes better sense than linear model intuitively, since the best wine should be more of a balance of all features. However, the accuray is not super great. Introducing more features or more complicated models might help to increase the accuracy.

------

# Final Plots and Summary

### Plot One
```{r, Plot_One}
ggplot(aes(x = sulphates), data = subset(redwines, sulphates < 1.2)) + 
  geom_histogram(binwidth = 0.005) +
  scale_x_log10(name = 'sulphates', breaks = seq(0,2,0.2)) +
  ggtitle("Distribution of Sulphates (x-axis log transformed)")
```

### Description One

The distribution of sulphates appears to be nomarl on log scale. This is fairly interesting. The reason may be that sulphates are more easily change exponentially.

### Plot Two
```{r, Plot_Two}
ggplot(redwines, aes(quality, volatile.acidity)) +
  geom_boxplot(aes(group=quality), outlier.shape = NA) +
  geom_jitter(alpha = 0.2, position = position_jitter(width = 0.4)) +
  scale_x_continuous(name = 'quality', breaks = seq(3,8,1)) +
  scale_y_continuous(name = 'volatile acidity') +
  ggtitle('Distribution of Volatile Acidity by Quality')
```

### Description Two

The relationship between volatile acidity and quality is significant according to the dataset. Usually, as volatile acidity decreases, quality increases. However, we can still see that wines with quality of 5 and 6 have many points in the same range of volatile acidity. Definitely other features should be concerned to help us to predict quality.

### Plot Three
```{r, Plot_Three}
ggplot(aes(x = sulphates, y = alcohol, color = quality.factor), data = redwines) +
  geom_point(alpha = 0.5, size = 1) +
  scale_x_continuous(name = 'sulphates') +
  scale_y_continuous(name = 'alcohol') +
  labs(colour = "quality") +
  scale_color_brewer(palette="YlOrRd") +
  ggtitle("Quality against Sulphates and Alcohol") +
  theme_dark()
```

### Description Three

The plot shows that alcohol and sulphates combined is helpful to distinguish quality of 5 and 7. Most wines with quality of 5 are in the bottom left in the plot, while most wines with quality of 7 are in the top right in the plot. However, we cannot easily tell quality 5 from quality 6 from this plot. Other features should be considered to help us.

------

# Reflection

Qualities of wines are rated subjectively by wine experts, and other features are created from objective tests. We want to build a model that is able to predict the quality from those objective features.

At first I studied the dataset and remove some duplicates to avoid . Secondly, each individual feature is investigated to see its pattern. I also checked their relationship with each other and quality. It turns out that no feature shows a strong linear relationship with quality. Therefore I turned to try tree models and check if some features combined could help to distinguish different qualities.

The tree model is actually doing good when predicting qualities. However, the predicting power of the tree model is not super great, and the model may be overfitting. Future works could be done on getting more data to train and test, and other more complicated models like random forest or SVM (super vector machine). I actually checked the paper linked to the dataset, where the authors achieved a promising model based on SVM. The paper also shows that the predicting power is stronger on the white wines dataset, which has more records than the red wines dataset.



# References

1. https://s3.amazonaws.com/content.udacity-data.com/courses/ud651/diamondsExample_2016-05.html
2. http://www.statmethods.net/advstats/cart.html
3. http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/
4. http://www.sciencedirect.com/science/article/pii/S0167923609001377